---
title: "ORA enrichment - single cell metabolomics"
author:
  - name : Bishoy Wadie
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ORA_enrichment-single_cell_metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Single-cell metabolomics generates output similar to scRNA-Seq, resulting in a matrix of metabolites (m) by cells (c), where values represent the abundance of each metabolite per cell. As with any single-cell analysis pipeline, enrichment analysis is typically one of the final steps to interpret the differential markers identified earlier.

Overrepresentation analysis (ORA) is the most common enrichment method for both bulk and single-cell datasets, regardless of the molecular readout. In metabolomics, several methods exist for metabolite enrichment, with [MetaboAnalyst](https://www.metaboanalyst.ca/) being the most popular. [MetaboAnalyst](https://www.metaboanalyst.ca/) is a robust platform offering various analyses for metabolomics data at different stages (raw, preprocessed, biomarkers, etc.). We recommend users explore this tool for additional analyses beyond enrichment.

For MS1-based metabolomics datasets, particularly in imaging MS, the inherent molecular ambiguity in metabolite identification complicates downstream analyses, including enrichment. To our knowledge, no other metabolite enrichment method addresses isomeric/isobaric ambiguity in enrichment analyses.

In this notebook, we demonstrate how to use `bmetenrichr` for ORA enrichment in single-cell metabolomics datasets while accounting for isomeric/isobaric ambiguity.

## Loading packages

```{r "load packages", message=FALSE}
library(bmetenrichr) #TODO change package name
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Dataset

The single-cell dataset used is from the SpaceM paper, which models NASH by stimulating Hepa-RG cells with fatty acids and other inhibitors compared to a control, followed by MALDI imaging MS.

The data is freely available in [MetaboLights](https://www.ebi.ac.uk/metabolights/editor/MTBLS78/descriptors). 

### Download single-cell matrices and associated metadata

`NASH_scm` contains the single-cell metabolite matrix which will be main input as well as condition per cell in `NASH_scm$metadata`. These are the main required files to run single-cell metabolomics enrichment. `condition_metadata` contains the [METASPACE](https://metaspace2020.eu/) dataset names for each replicate, while `metaspace_annotations` contains the annotation results for each dataset in the [SpaceM](https://metaspace2020.eu/project/Rappez_2021_SpaceM?tab=datasets) project on METASPACE. We will use the annotation results and corresponding FDR thresholds to select metabolites as input query and corresponding universe for enrichment.

```{r "Download data", eval=FALSE}
NASH_scm_tmp = tempfile()
download.file("path/to/RDS", destfile = NASH_scm_tmp)
NASH_scm = readRDS(NASH_scm_tmp)

condition_metadata_tmp = tempfile()
download.file("/path/to/RDS", destfile = condition_metadata_tmp)
condition_metadata = readRDS(condition_metadata_tmp)[["metaspace_dataset_names"]]

metaspace_annotations_tmp = tempfile()
download.file("/path/to/RDS", destfile = metaspace_annotations_tmp)
metaspace_annotations = readRDS(metaspace_annotations)

```

### Prepare the input data

```{r "Prepare data", eval=FALSE}
scm = NASH_scm$scm %>% 
  as.matrix() %>% 
  t()
conds = NASH_scm$metadata %>%
  column_to_rownames("Cell")
conds = conds[colnames(scm),]

conds_unique = conds %>% 
  dplyr::distinct()

metaspace_annotations = metaspace_annotations %>% 
  dplyr::left_join(condition_metadata, by = c("ds_name" = "dataset_name")) %>%
  dplyr::rename("Replicate" = "Condition") %>%
  dplyr::left_join(conds_unique)

```

### Filter metabolites and specify conditions

Here we specify the reference and query conditions as `cond_x` and `cond_y`, respectively. And since METASPACE provides FDR-controlled annotations, we will select annotations passing `desired_fdr` as query and all detected annotations for a given annotation database (`desired_annot_db`) as custom universe.

```{r "Filter metabolites", eval=FALSE}
cond_x = "U"
cond_y = "F"
desired_fdr = 0.1
desired_annot_db = "HMDB"

annots_des_fdr = metaspace_annotations %>%
  dplyr::filter(Condition %in% c(cond_x, cond_y),
                fdr <= desired_fdr,
                str_detect(db, desired_annot_db)) %>%
  pull(formula_adduct) %>%
  intersect(rownames(scm))

custom_univ = metaspace_annotations %>%
  dplyr::filter(Condition %in% c(cond_x, cond_y),
                str_detect(db, desired_annot_db)) %>%
  pull(formula_adduct) %>%
  intersect(rownames(scm))

input_scm = scm[annots_des_fdr,]

```

## Initialize enrichment object

The first step is creating a `bmetenrichr` object which contains the input matrix in `scmatrix` and the conditions for each cell specified in `conditions`. We need to specify the `enrichment_type` as "ORA". `annot_db` corresponds to the annotation database used during the annotation, more relevant if annotation is performed using [METASPACE](https://metaspace2020.eu/). The databases supported are ("CoreMetabolome", "HMDB","SwissLipids","LipidMaps"), if you want to provide a custom annotation database, you can provide it using `annot_custom_db` argument. Here we will use "HMDB" as annotation database.

Since we want to consider isomeric/isobaric ambiguity, we set either `consider_isomers` or `consider_isobars` to `TRUE`. If both were set to `FALSE`, it will run classic ORA with no bootstrapping. We also specify polarization_mode to `positive` since these datasets were acquired in `positive` mode.

For enrichment background, we use the `background_type` argument to select one of the possible background types :

1.  `LION` : Uses [LION ontology](http://www.lipidontology.com/). Only for Lipids.
2.  The following types are curated from [RAMP-DB 2.0](https://pubmed.ncbi.nlm.nih.gov/36373969/) :
    1.  `super_class` : Most coarse-grained classification
    2.  `main_class` : Fine-grained sub classification compared to `super_class`
    3.  `sub_class` : Most fine_grained sub classification.
    4.  `pathways` : Biological pathways curated from KEGG, HMDB, Reactome and WikiPathways.

We also specify the background `molecule_type` by specifying either `Metabo` for metabolites or `Lipid` for lipids. To pull the relevant background which is internally built in `initEnrichment`, you can provide the previous arguments to the `Load_background` function to get list of terms and their associated molecules as follows :

```{r "Load Background", eval=FALSE}
bg = Load_background(mol_type = "Metabo",
                     bg_type = "sub_class", 
                     feature_type = "name")

```

Finally we specify `condition.x` and `condition.y` as reference and query conditions, respectively. While running ORA in `Run_enrichment`, fold changes will be computed in `condition.y` relative to `condition.x`.

```{r "init enrich object", eval=FALSE}
ORA_boot_obj = initEnrichment(scmatrix = input_scm, conditions = conds$Condition,
                     enrichment_type = "ORA",annot_db = "HMDB",
                     consider_isomers = T, consider_isobars = T,
                     polarization_mode = "positive",
                     background_type = "sub_class",
                     molecule_type = "Metabo",
                     condition.x = cond_x,
                     condition.y = cond_y)
```

There are additional arguments to `initEnrichment` that could be used depending on your dataset metadata and any other prior knowledge. Check documentation of `?initEnrichment` for more information on these arguments.

## Running ORA bootstrapping-based enrichment

Based on the initialized object, we provide `Run_enrichment` function as a wrapper around the main enrichment functions :

1.  `Run_bootstrap_ORA()`

2.  `Run_simple_ORA()`

3.  `Run_bootstrap_MSEA()`

4.  `Run_simple_MSEA()`

So you can also provide additional arguments to `Run_enrichment` from the argument list of the relevant function from the above list.

In this example, `Run_enrichment` will first calculate log fold change to separate metabolites into upregulated and downregulated if `Run_DE` is set to `FALSE`. If it was set to `TRUE`, the function `seurat_WilcoxDETest` (a wrapper for [WilcoxDETest](https://www.rdocumentation.org/packages/Seurat/versions/2.3.4/topics/WilcoxDETest) from Seurat) will be used to calculate p-values based on a wilcoxon rank-sum test which will be used in addition to the previously computed log fold changes to select the markers for ORA based on the `DE_pval_cutoff` and `DE_LFC_cutoff` arguments in the `Run_enrichment` function.

The `min.pct.diff` argument of 0.1 specifies that a marker must have at least a 10% difference in detection between cells in both conditions to be considered differentially abundant.

From the additional list of arguments to `Run_bootstrap_ORA` we recommend defining the following arguments to `Run_enrichment` :

-   `custom_universe` : List of background metabolites to be used. More specifically, these represent the set of all possible metabolites that were measured in a given dataset (optionally under a given threshold). If not provided, all metabolites in the selected background will be used and it might lead to potentially misleading results.

-   `report_ambiguity_scores` : Useful to understand the degree of isomeric/isobaric ambiguity per metabolite.

-   `n_bootstraps`: Number of bootstrap iterations. The default is 50, but increasing this number generally improves accuracy, though it will slow down the process. Adjust according to your needs. We recommend a minimum of 50 and a maximum of 1000. 100 is acceptable.

```{r "Run ORA boot", eval = FALSE}
ORA_boot_res = Run_enrichment(object = ORA_obj,
                         custom_universe = custom_univ,
                         report_ambiguity_scores = T,
                         DE_LFC_cutoff = 0,min.pct.diff = 0)

```

## Output

The main output of `Run_bootstrap_ORA` is a list of 2 dataframes :

1.  "unfiltered_enrich_res" (`ORA_boot_res[["unfiltered_enrich_res"]]`) : Data frame containing the enrichment results for each term and bootstrap and the contingency table used for ORA.

2.  "clean_enrich_res" (`ORA_boot_res[["clean_enrich_res"]]`): Summary statistics per term passing the filters specified in `Run_bootstrap_ORA`

Check `?Run_bootstrap_ORA` for more information on the output and other parameters.

Since we used `Run_enrichment` as a wrapper around `Run_bootstrap_ORA` we get the ORA results separately for **upregulated**, **downregulated**, and **all** metabolites based on the calculated `DE_LFC_cutoff` specified above. If you use `Run_bootstrap_ORA` directly, you can provide a list of multiple conditions as input directly as long as you know the markers *a priori*.

To get a summary of the terms passing each filter, we can call `passed_filters_per_term` function on any unfiltered dataframe and any combinations of filters to understand why a given term was excluded in the final summarized results. Check `?passed_filters_per_term` for more information on the filters.

```{r "Summary boot", eval=FALSE}
enrich_ORA_summary = passed_filters_per_term(unfiltered_df = ORA_boot_res$upregulated$unfiltered_enrich_res,
                        enrich_type = "ORA", min_intersection = 3,alpha_cutoff = 0.05,q.val_cutoff = 0.2,boot_fract_cutoff = 0.5)
enrich_ORA_summary = enrich_ORA_summary[order(enrich_ORA_summary$pass_all_filts, decreasing = T),]
head(enrich_ORA_summary)
```

## Visualization

### Barplot

```{r Bar_plot, message=FALSE, eval=FALSE, echo=T}
p = barplot_ORA_boot(ORA_boot_res = ORA_boot_res,collapse_multi_cond = T)
```

### Dotplot

```{r dotplot, message=FALSE, eval=FALSE}
multi_cond_collapse = collapse_ORA_boot_multi_cond(ORA_boot_res_list = ORA_boot_res)
dotplot_ORA(ORA_res = multi_cond_collapse$clean_enrich_res)
```

### Ridge plots

To compare the distribution of term/query overlap size across bootstraps we can plot the distribution of the overlap size across terms of interest. Here we use `ridge_bootstraps` function to plot the distribution of the terms enriched in both upregulated and downregulated markers found in `multi_cond_collapse` using the **upregulated** results only.

```{r "Ridge", eval=FALSE}
ridge_bootstraps(enrich_res = multi_cond_collapse$unfiltered_enrich_res,
                 terms_of_interest = c(multi_cond_collapse$clean_enrich_res$Term),
                 condition = "upregulated")
```

### Comparative distribution of marker ions

We also provide a simple function to get the TP markers associated with a given term which correspond to the input ion in the `input_scm` matrix used as input.

```{r "Get markers", eval=FALSE}
TP_ions = get_TP_markers_per_Term(ORA_boot_df = ORA_boot_res$downregulated$unfiltered_enrich_res,
                                  term_of_interest = "Glycerophosphocholines")
TP_ions = map_TP_markers_to_ions(markers = TP_ions,
                                 scm_ions = rownames(input_scm))
TP_ions
```

We can then select an ion of interest to check distribution of intensities in the specified conditions using the `compare_metabo_distr` function which takes the `ORA_boot_obj` as input, the ions and the conditions of interest.

```{r "plot_compar_dist", eval=FALSE}
compare_metabo_distr(ORA_boot_obj, metabolite = TP_ions[5],
                     conds_of_interest = c(cond_x, cond_y))
```
